\section{Background and Motivation} \label{sec:back}

This section provides relevant background for the $\mathcal{CE}$ machine,
outlining lambda calculus, evaluation strategies, and Curien's calculus of
closures.

\subsection{Preliminaries}

We begin with the simple lambda calculus:  $$ t::= x \; | \;  \lambda x.t \; |
\;  t \; t $$ where $x$ is a variable, $\lambda x.t$ is an abstraction, and $t
\; t$ is an application. We also use lambda calculus with deBruijn indices,
which replace variables with a natural number indexing into the binding lambdas.
This calculus is given by the syntax: $$ t::= i \; | \; \lambda t \; | \; t \; t
$$ where $i \in \mathbb{N}$. In both cases, we use the standard Barendregt
syntax conventions, namely that applications are left associative and the bodies
of abstractions extend as far as possible to the right
~\cite{barendregt1984lambda}.  A \emph{value} in lambda calculus refers to an
abstraction. We are concerned only with evaluation to weak head normal form
(WHNF), in which evaluation terminates on an abstraction without entering its
body.

In mechanical evaluation of expressions, it would be too inefficient to perform
explicit substitution. To solve this, the standard approach uses closures
~\cite{landin1964mechanical,curien1991abstract,jonesstg,biernacka2007concrete}.
Closures combine a term with an environment, which binds the free variables in
the term to closures. 

For a formal basis, we use Curien's calculus of closures, given in
Figure~\ref{fig:calcclos}~\cite{curien1991abstract}.  It is a formalization of
closures with an environment represented as a list of closures, indexed by
deBruijn indices. We will occasionally modify this calculus by replacing the
deBruijn indices with variables for readability, in which case variables are
looked up in the environment instead of indexed, e.g., $t[x = c, y = c'])$
~\cite{barendregt1984lambda}. We also add superscript and subscript markers to
denote unique syntax elements, e.g., $t', t_1 \in \textnormal{Term}$. 

\subsection{Evaluation Strategies} \label{sec:eval}

There are three evaluation strategies for lambda calculus worth noting:
call-by-value, call-by-need, and call-by-name.  Call-by-value evaluates every argument
to a value, whereas call-by-need and call-by-name only evaluate an argument if
it is needed.  If an argument is needed more than once, call-by-name re-computes
the value, where call-by-need memoizes the value, so it is computed at most once.
Thus, call-by-need attempts to embody the best of both worlds---never repeat
work (call-by-value), and never perform unnecessary work (call-by-name). These
are intuitively \emph{good} properties to have, and we illustrate the
correctness of such an intuition with the following example, modified from
~\cite{danvy2013synthetic}:

$$ \overbrace{c_m (c_m (\cdots(c_m}^{m} \; id \;  id)\cdots) id) \; true \; id
\; bottom $$ where $c_n = \lambda s.\lambda z.\overbrace{s \; (s \cdots (s}^{n}
\; z) \cdots) $, $true = \lambda t.\lambda f.t$, $id=\lambda x.x$, and $bottom =
(\lambda x.x \; x) \lambda x.x \; x$. Call-by-value never terminates,
call-by-name takes exponential time, and call-by-need takes only polynomial time
~\cite{danvy2013synthetic}. Of course, this is a contrived example, but it
illustrates why call-by-need has some desirable properties.

In practice, however there are significant issues with call-by-need evaluation.
We focus on the following issue: \emph{Delaying a computation is slower than
performing it immediately.} This issue is well known in the literature, and has
become part of the motivation for a tool called \emph{strictness analysis},
which transforms non-strict evaluation to strict when possible
\cite{mycroft1982abstract,wadler1987projections}. 

\subsection{Existing Call-by-Need Machines}

Diehl et al. ~\cite{diehl2000abstract} review the call-by-need
literature in detail.  Here we summarizes a few key points.

The best known machine for lazy evaluation is the Spineless Tagless
G-Machine (STG machine), which underlies the Glasgow Haskell Compiler (GHC). 
STG uses flat environments that can be allocated on the stack, the heap,
or some combination ~\cite{jonesstg}. The STG machine is the
most widely used lazy evaluation machine. 

Two other influential lazy evaluation machines relevant to the $\mathcal{CE}$
machine are: the call-by-need Krivine machines
~\cite{lkm,krivine2007call,sestoft}, and the three instruction machine (TIM)
~\cite{TIM}.  Krivine started as an approach to call-by-name evaluation, and was
later extended to call-by-need
~\cite{krivine2007call,sestoft,danvy2013synthetic,lkm}.  The $\mathcal{CE}$
modifies the lazy Krivine machine to capture the environment sharing given by
the cactus environment. The TIM is an influential
implementation of call-by-need and call-by-name ~\cite{TIM}.  It involves, as
the name suggests, three machine instructions, \texttt{TAKE}, \texttt{PUSH}, and
\texttt{ENTER}. In Section~\ref{sec:impl}, we follow Sestoft ~\cite{sestoft} and
re-appropriate these instructions for the $\mathcal{CE}$ machine.

There has also been recent interest in \emph{heapless} abstract
machines for lazy evaluation. Danvy et al. ~\cite{danvy2012inter} and
Garcia et al.  ~\cite{garcia2009lazy}, independently derived similar
machines from the call-by-need lambda calculus
~\cite{ariola1995call}. We will return to this work in
Section~\ref{sec:disc}.  These are interesting approaches, but it is
not yet clear how these machines could be implemented efficiently.
