\section{Environment Representations} \label{sec:env}

As mentioned in Section~\ref{sec:back}, environments bind free variables to
closures. There is significant flexibility in how they can be represented. In
this section we review this design space in the context of existing work, both
for call by value and call by need.\footnote{Some work refers to this
space as \emph{closure} representation rather than \emph{environment}
representation~\cite{shao1994space,appel1988optimizing}.  Because the term
part of the closure is simply a code pointer and the
interesting design choices are in the environment, we refer to
the topic as environment representation.}

There are two common approaches that span the design space for environment
representation: \emph{flat} environments and \emph{shared} environments (also
known as linked environments)~\cite{appel1988optimizing,shao1994space}. A flat
environment is one in which each closure has its own record of the terms 
its free variables are bound to. A shared environment is one in which parts of
that record are shared among multiple
closures~\cite{appel1988optimizing,shao1994space}. For example, consider the
following term: $$(\lambda x.(\lambda y.t) (\lambda z.t_1)) t_2$$ Assuming the
term $t$ has both $x$ and $y$ as free
variables, we must evaluate it in the environment binding both $x$ and $y$.
Similarly, assuming $t'$ contains both $z$ and $x$ as free variables, we must
evaluate it in an environment containing bindings for both $x$ and $z$. Thus, we
can represent the closure for evaluating $t$ and $t_1$  as $$t[x=t_2[\bullet],
y=c]$$ and $$t_1[x=t_2[\bullet], z=c_1]$$ respectively.  These are examples of
\emph{flat} environments, where each closure comes with its own record of all of
its free variables. Because of the nested scope of the given term, $x$ is bound
to the same closure in each environment.  Thus, we can also create a shared,
linked environment, represented by the following diagram:

\begin{center}
\begin{tikzpicture}[ 
  edge from parent path={(\tikzchildnode\tikzchildanchor) edge [-latex] (\tikzparentnode\tikzparentanchor)},
  level distance=1cm
]
\node (d) {$\bullet$} child{node (a) {$x=t_2[\bullet]$} child{node (b) {$y=c$}} child{node (c)
{$z=c_1$}}};

\end{tikzpicture}
\end{center}
Now each of the environments is represented by a linked list, with the binding
of $x$ shared between them. This is an example of a \emph{shared} environment
~\cite{appel1988optimizing}. This shared, linked structure dates back to the 
first machine for evaluating expressions: Landin's SECD
machine~\cite{landin1964mechanical}.

The drawbacks and advantages of each approach are well known. With a flat
environment, variable lookup can be performed with a simple offset
~\cite{jonesstg,appel2006compiling}. On the other hand, significant
duplication can occur, as we will discuss in Section~\ref{sec:exist}.
With a shared environment, that duplication is removed, but at the cost of
possible link traversal upon dereference. 

As with most topics in compilers and abstract machines, the design space is
actually more complex. For example, Appel and Jim show a wide range of hybrids
~\cite{appel1988optimizing} between the two, and Appel and Shao
~\cite{shao1994space} show an optimized hybrid that aims to achieve the benefits
of both approaches. And as shown in the next section, choice of evaluation
strategy further complicates the picture.

\subsection{Existing Call-by-Need Environments} \label{sec:exist}

Existing call by need machines use flat environments with a heap of
closures~\cite{jonesstg,TIM,johnsson1984efficient,boquist1997grin}. These
environments may hold some combination of primitive values and pointers into the
heap. The pointers and heap implement the memoization of results required for
call by need. Returning to the earlier example, $(\lambda x.(\lambda y.t)
(\lambda z.t_1)) t_2$, we can visualize the execution state when entering $t$ as
follows:

\begin{center}
\textbf{Closure}
\begin{align*}
t[x=p, y=p_1] \\
\end{align*}
\textbf{Heap}
\begin{align*}
p &\mapsto t_2[\bullet] \\
p_1 &\mapsto \lambda z.t_1[x=p] 
\end{align*}
\end{center}

If $t_2[\bullet]$ is not in WHNF (this sort of unevaluated closure is called a
\emph{thunk}~\cite{ingerman1961way,peyton1992implementing}), then if it is
entered in either the evaluation of $t$ or $t_1$, the resulting value will
overwrite the closure at $p$. The result of the computation is then shared with
all other instances of $x$ in $t$ and $t_1$. 

To better understand one of the costs of flat environments, consider the
following example: $$\lambda g.\lambda x.\lambda y.\lambda z.g
\; t \;t'$$ Without any preprocessing, the standard cost of duplication applies:
if $x$, $y$, and $z$ are free in $t$ and $t'$, then $t$ and $t'$ will have
large, identical environments and their closures will be expensive to create.
Compile-time transformation ~\cite{peyton1992implementing} (tupling arguments)
helps, but it is not clear that the technique extends to the general case. We
claim that the machine itself should avoid duplicating environments.

Depending on $g$, either or both of the closures created for its arguments may
not be evaluated.  Therefore, it is possible that the work of creating the
environment for that thunk will be wasted. This waste is well known, and
existing approaches address it by avoiding thunks as much as possible
~\cite{jonesstg,johnsson1984efficient}. Unfortunately, in cases like the above
example, this is impossible; the thunks are necessary. Indeed, even if we tuple
the closures for $x$, $y$, and $z$, that tuple could be wasted if neither $t$ nor
$t'$ is used. Thus, we would like to minimize the cost of creating such thunks.

Thunks are special in another way.  Recall the standard advantage of the flat
environment: quick variable lookups. In a lazy language, this advantage is
lessened by the following fact: \emph{a thunk can only be entered once}. After
it is entered, it will be overwritten with a value, so the next time that heap
location is entered it will be entered with a value and a different environment.
Thus, the work to ensure that the variable lookup is fast will be used only
once. This is in contrast to a call by value language, in which every closure is
constructed for a value, which can be entered an arbitrary number of times. 

A more subtle drawback of the flat environment representation is that
environments can vary in size, and thus a value in WHNF can be too large to fit
in the space allocated for the thunk it is replacing. This problem is discussed
in~\cite{jonesstg}, where the proposed solution is to put the value closure in
a fresh location in the heap where there is sufficient room. The original
thunk location is then replaced with an indirection to the value at the freshly
allocated location. These indirections are removed during garbage collection,
but do impose some cost, both in runtime efficiency and implementation
complexity~\cite{jonesstg}.

We have thus far ignored a number of details with regard to current
implementations. For example, the STG machine can split the flat environment, so
that part is allocated on the stack and part on the heap.  The TIM allocates its
flat environments separately from its closures so that each closure is a code
pointer, environment pointer pair~\cite{TIM}. while the STG machine keeps
environment and code co-located~\cite{jonesstg}. Still, the basic design
principle holds: a flat environment for each closure allows quick variable
indexing, but at the costs given above.

An intuitive view of the flat environment representation in a call by need
language is that whenever a term might be needed, the necessary environment is
constructed from the current environment.  This operation is expensive, and
it is wasted if the variable is never entered. This potentially wasted cost is
one of the primary motivations for the current work: \emph{lightweight
closure creation}. In other words, we would like to minimize the cost of saving
the current environment.

See Figure~\ref{fig:designspace} for a simple summary of the design space
relevant to this paper. There are existing call by value machines with both flat
and shared environments, and call by need machines with flat environments. As
far as we are aware, we are the first to use a shared environment to implement
lazy evaluation.

\begin{figure}
\begin{tabularx}{\textwidth}{l | X | X}
                & Flat Environment     & Shared Environment \\ \hline
  Call by need  & STG, TIM, GRIN & $\mathcal{CE}$ Machine (this paper) \\
  Call by value & ZAM, SML & ZAM, SECD, SML \\
\end{tabularx}
\caption{Evaluation strategy and environment structure design space. Each
acronym refers to an existing implementation. Some implementations use multiple
environment representations.}
\label{fig:designspace}
\end{figure}

