\section{Performance Evaluation} \label{sec:eval}

This section describes experiments that asses the strengths and weaknesses of
the $\mathcal{CE}$ machine. We evaluate using benchmarks from the \texttt{nofib}
benchmark suite. Because we have implemented only machine integers, and must
translate the examples by hand, we use a subset of the \texttt{nofib} suite that
does not use floating point values or arrays. Ideally, we would have a compiler
from Haskell to our intermediate representation. We discuss the possibility of a
Haskell compiler using our implementation as a backend in
Section~\ref{sec:disc}. A list of the benchmarks used and a brief description is
given in Figure~\ref{fig:bench}.

\begin{figure}
\begin{itemize}
\item \textbf{exp3:} A peano arithmetic benchmark. It computes $3^8$ in
peano arithmetic, and prints the result. 
\item \textbf{queens:} Computes the number of solutions to the nqueens problem
for an n by n board.
\item \textbf{primes:} A simple primes sieve that computes the nth prime.
\item \textbf{digits-of-e1:} A calculation of the digits of $e$ using continued
fractions. Computes the first n digits of $e$.
\item \textbf{digits-of-e2:} Another calculation of the digits of $e$ using an
infinite series. Computes the first n digits of $e$. 
\item \textbf{fib:} Naively computes the n'th Fibonacci number.
\item \textbf{fannkuch:} A counting benchmark counting the number of reverses of
a subset of a list.
\item \textbf{tak:} A synthetic benchmark involving basic recursion.
\item \textbf{expexp:} Our own synthetic benchmark, using church numerals to
calculate $3^8-3^8$. We add this example to compare implementations on
equivalent pure lambda terms.
\end{itemize}
\caption{Description of Benchmarks}
\label{fig:bench}
\end{figure}

We compare the $\mathcal{CE}$ machine to two existing implementations:

\begin{itemize}
\item GHC: The Glasgow Haskell compiler. A high performance, optimizing compiler
based on the STG machine \cite{jonesstg}.
\item UHC: The Utrecht Haskell compiler. An optimizing compiler based on the
GRIN machine \cite{boquist1997grin, dijkstra2009architecture}.
\end{itemize}

We use GHC version 7.8.3 and UHC version 1.1.5. We compile with both -O0 and
-O3, and show the results for both. For GHC, we use the -H flag from the runtime
system options to set the default heap size to the same value as the
$\mathcal{CE}$ machine (1G) to avoid the measuring the performance of GC
implementations. The tests were run on an Intel(R) Xeon(R) CPU E5-4650L @
2.60GHz, running Linux version 3.16. 
