\documentclass[preprint]{sigplanconf}

\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{natbib}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{mathtools}
\usetikzlibrary{chains,fit,shapes,calc}
\usepackage{verbatim}
\usepackage{semantic}
\usepackage{tabu}
\usepackage{amsthm}
\usepackage{mathptmx}
\usepackage{todonotes}
\newcommand{\concat}{\ensuremath{+\!\!\!\!+\,}}  

\newtheorem{prop}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{defin}{Definition}

\begin{document}

\conferenceinfo{ICFP '14}{September 1-3, 2014, Gothenburg, Sweden} 
\copyrightyear{2014} 
\copyrightdata{[to be supplied]} 

\titlebanner{Preprint}        % These are ignored unless
\preprintfooter{Cactus Environment Machine}   % 'preprint' option specified.

\title{Cactus Environment Machine}
\subtitle{Shared Environment Call by Need}

\authorinfo{George Stelle}
           {University of New Mexico}
           {stelleg@cs.unm.edu}
\authorinfo{Darko Stefanovic}
           {University of New Mexico}
           {darko@cs.unm.edu}
\authorinfo{Stephen L. Olivier}
           {Sandia National Laboratories}
           {slolivi@sandia.gov}
\authorinfo{Stephanie Forrest}
           {University of New Mexico}
           {forrest@cs.unm.edu}

\maketitle

\begin{abstract}
Existing machines for lazy evaluation use a \emph{flat} representation of
environments, storing the terms associated with free variables in an array.
Combined with a heap, this structure supports the shared intermediate
results required by lazy evaluation.  This paper proposes and describes an
alternative approach that uses a \emph{shared} environment to implement lazy
evaluation.  We show how a shared environment can act as both an environment and
a mechanism for sharing results. To formalize this approach, we describe a novel
calculus that makes the shared environment explicit, as well as a machine to
implement the calculus: the \emph{Cactus Environment Machine} ($\mathcal{CE}$
machine). A simple compiler implements the machine and is used to run
experiments for assessing performance.  The results show reasonable performance
and suggest that with the addition of some common optimizations, the
$\mathcal{CE}$ machine could be competitive with existing high performance lazy
evaluation implementations.
%In addition, we implement a simple compiler that targets native
%code and show that it has reasonable performance despite some significant
%but addressable shortcomings.
\end{abstract}

\input{intro}
\input{background}
\input{env}
\input{calc}
\input{mach}
\input{implement}
\input{evaluation}
\input{results}
\input{discussion}
\input{conclusion}

\section{Acknowledgments}
Sandia is a multiprogram laboratory operated by Sandia Corporation, a Lockheed Martin Company, for the United States Department of Energy’s National Nuclear Security Administration under contract DE-AC04-94AL85000.

% We recommend abbrvnat bibliography style.
\bibliographystyle{abbrvnat}
\bibliography{annotated}

\input{appendix}
\end{document}
